Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2025-01-27 CSEDU2025/Paper.tex

Recent advances in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \citep{Peng23}
\citep{Pandey24} and education \citep{Puryear22}. In programming courses,
students can employ GenAI tools to improve their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists in
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \citep{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and try to foster a deeper understanding of programming concepts
\citep{Labadze23}.

The study conducted by \cite{chan23} revealed that both undergraduate and
postgraduate students have positive attitudes towards the use of GenAI.
A systematic review organized by \cite{Lo24} demonstrated that students can
effectively learn from ChatGPT, resulting in improved comprehension and academic
achievement \citep{Callejo24}. Furthermore, it was observed that ChatGPT may
allow students to regulate their learning pace \citep{Baha24} and can
potentially support self-regulated learning, particularly for students with
prior technical and disciplinary knowledge \citep{Xia23}.

@@

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.1} % Increase the spacing between rows
    \begin{tabular}{p{3.8cm} p{6cm} p{4.5cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Example of prompt} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \textit{Would this code be ok? \{code\}} \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. & \textit{What does it mean for a method to be static?} \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. & \textit{The correct answer is B} \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. & \textit{A class inherits from another write this code} \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. & \textit{Give me the code for a selection sort} \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). & \textit{A heap is a data structure appropriate for: \{options\}} \\
      Language Change (LC) & Prompts that request a change of idiom. & \textit{In Spanish} \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. & \textit{Thanks} \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

\subsection{Metacognition Related Works}

Learning based on metacognitive strategies has been shown to be effective in
enhancing students’ performance and skills. \cite{Zheng19} investigated the
effects of metacognitive scaffolds on group behavior, performance, and cognitive
load in computer-supported collaborative environments. The results showed
positive impacts on metacognitive behavior and group performance without
increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
collaborative approach to enhance performance in collaborative programming. The
findings indicated that this approach influenced computational thinking,
critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
explored how metacognitive instruction affects computational thinking, critical
thinking, and metacognitive skills in collaborative programming, observing

@@

\cite{Sidra24} explored the importance of metacognitive skills in an AI-enabled
workforce, focusing on how AI, particularly tools using Large Language Models,
affects human decision-making. They highlighted challenges arising from AI's
lack of human-like metacognition and identified key differences between human
AI and human-human collaboration. Using dual-process theory, they emphasized the
need for workers to develop metacognitive skills like planning, monitoring, and
evaluating to counter cognitive biases and enhance decision-making. The study
recommended improving AI design and training to boost workers' metacognitive
abilities for more effective collaboration.

@@

The analysis phase involved using Python/Pandas scripts to extract information
from previously classified data. The data sets, and code used in this process
are available at https://github.com/rodrigoprestesmachado/uc3m.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.61]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.41]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.1} % Adjust the spacing between rows
    \begin{tabular}{p{4cm} p{3cm} p{3cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.34 & 1.95 \\
      Uncategorized & 1.70 & 3.59 \\
      Language Change & 2.23 & 3.71 \\
      Multiple Question & 2.47 & 5.66 \\
      \hdashline
      \textit{Overall} & \textit{3.78} & \textit{6.95} \\
      \hdashline
      Student Correction & 4.07 & 8.42 \\
      Debugging Help & 5.30 & 10.80 \\
      Conceptual Questions & 5.47 & 10.46 \\
      Complete Solution & 7.65 & 11.03 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

There are ways to deepen this study; based on previous references, 60 minutes
was the standard spacing measure. However, further studies will deepen this
analysis using a quantitative variable to better represent different spacing
intervals. Future research can also include additional metacognitive strategies,
which will involve interpreting user messages in depth. For instance, the
Monitoring strategy was evident when students expressed that a concept remained
unclear and requested further clarification. Another necessary analysis involves
investigating the relationship between student profiles, metacognitive
strategies for using GenAI, and learning outcomes.

>>>>

Recent advances in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming and education \citep{Puryear22}.
In programming courses, students can employ GenAI tools to improve their
understanding, receive personalized feedback, and access detailed explanations.
For example, GitHub Copilot, a tool that is integrated into development
environments, assists in providing real-time suggestions and accelerating code
writing, which could help streamline the learning process. Educational chatbots,
on the other hand, try to guide students through coding challenges, answer
questions, and try to foster a deeper understanding of programming concepts.

The study conducted by \cite{chan23} revealed that both undergraduate and
postgraduate students have positive attitudes towards the use of GenAI.
A systematic review organized by \cite{Lo24} demonstrated that students can
effectively learn from ChatGPT, resulting in improved comprehension and academic
achievement \citep{Callejo24}. Furthermore, it was observed that ChatGPT may
allow students to regulate their learning pace and can potentially support
self-regulated learning, particularly for students with prior technical and
disciplinary knowledge \citep{Xia23}.

@@

\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1} % Increase the spacing between rows
    \begin{tabular}{p{3.8cm} p{6cm} p{4.5cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} & \textbf{Example of prompt} \\
      \hline
      Debugging Help (DH) & Prompts that seek help to identify, fix errors, or understand the provided code snippet. & \textit{Would this code be ok? \{code\}} \\
      Conceptual Question (CQ) & Prompts that are more about understanding concepts than specific code. & \textit{What does it mean for a method to be static?} \\
      Student Correction (SC) & Prompts where the student corrects the chatbot. & \textit{The correct answer is B} \\
      Code Snippet  (CS) & Prompts that ask for a specific part of the code, like a function or a segment. & \textit{A class inherits from another write this code} \\
      Complete Solution (CSO) & Prompts that request an entire solution or a complete code snippet. & \textit{Give me the code for a selection sort} \\
      Multiple Question (MQ) & Prompts where the user wants to solve a multiple choice exercise (Quiz). & \textit{A heap is a data structure appropriate for: \{options\}} \\
      Language Change (LC) & Prompts that request a change of idiom. & \textit{In Spanish} \\
      Uncategorized (U) & Prompts that do not fit into any of the above categories. & \textit{Thanks} \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

% \subsection{Metacognition Related Works}

% Learning based on metacognitive strategies has been shown to be effective in
% enhancing students’ performance and skills. \cite{Zheng19} investigated the
% effects of metacognitive scaffolds on group behavior, performance, and cognitive
% load in computer-supported collaborative environments. The results showed
% positive impacts on metacognitive behavior and group performance without
% increasing the cognitive load. \cite{LiWei23} proposed a metacognition-based
% collaborative approach to enhance performance in collaborative programming. The
% findings indicated that this approach influenced computational thinking,
% critical thinking, and metacognitive awareness. Similarly, \cite{Wang23}
% explored how metacognitive instruction affects computational thinking, critical
% thinking, and metacognitive skills in collaborative programming, observing

@@

% \cite{Sidra24} explored the importance of metacognitive skills in an AI-enabled
% workforce, focusing on how AI, particularly tools using Large Language Models,
% affects human decision-making. They highlighted challenges arising from AI's
% lack of human-like metacognition and identified key differences between human
% AI and human-human collaboration. Using dual-process theory, they emphasized the
% need for workers to develop metacognitive skills like planning, monitoring, and
% evaluating to counter cognitive biases and enhance decision-making. The study
% recommended improving AI design and training to boost workers' metacognitive
% abilities for more effective collaboration.

@@

The analysis phase involved using Python/Pandas scripts to extract information
from previously classified data.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.6]{img/figure1.png}
  \caption{Categorization of messages.}
  \label{fig:graph1}
\end{figure}

\begin{figure*}[htbp]
  \centering
  \includegraphics[scale=0.39]{img/figure2.png}
  \caption{Examples of conversations.}
  \label{fig:graph2}
\end{figure*}

@@

\begin{table*}[htbp]
  \caption{Average and Standard Deviation of Different Categories (in minutes)}
  \begin{center}
    \renewcommand{\arraystretch}{1.0} % Adjust the spacing between rows
    \begin{tabular}{p{4cm} p{3cm} p{3cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Average} & \textbf{Standard Deviation} \\
      \hline
      Code Snippet & 1.34 & 1.95 \\
      Uncategorized & 1.70 & 3.59 \\
      Language Change & 2.23 & 3.71 \\
      Multiple Question & 2.47 & 5.66 \\
      \hdashline
      \textit{Overall} & \textit{3.78} & \textit{6.95} \\
      \hdashline
      Student Correction & 4.07 & 8.42 \\
      Debugging Help & 5.30 & 10.80 \\
      Conceptual Questions & 5.47 & 10.46 \\
      Complete Solution & 7.65 & 11.03 \\
      \hline
    \end{tabular}
    \label{tab:averages}
  \end{center}
\end{table*}

@@

There are ways to deepen this study. While 60 minutes was the standard spacing
measure, future research will explore different intervals using quantitative
variables. Additional metacognitive strategies, such as the Monitoring
strategy—where students seek clarification on unclear concepts—will also be
examined. Moreover, investigating the relationship between student profiles,
GenAI metacognitive strategies, and learning outcomes is essential.
